---
title: "01_make_data_.Rmd"
author: "Abisai Lujan"
date: "2025-01-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
https://github.com/rfordatascience/tidytuesday/blob/main/data/2022/2022-06-14/readme.md 

```{r, eval = FALSE, include=FALSE}
library("tidyverse")
```


```{r, eval = FALSE, include = FALSE}
drought <- read.csv("/Users/abisailujan/Documents/GitHub/DATA481-Breast_Cancer/Abisai")
head(drought)
tail(drought)
```



```{r, eval = FALSE, include = FALSE}
drought_fips <- read.csv("drought-fips.csv")
(drought_fips)
unique(drought_fips$State)
```

```{r, eval = FALSE, include = FALSE}
(names(drought_fips))
(names(drought))
```





```{r, eval = FALSE, include = FALSE}
(min(drought_fips$DSCI))
(max(drought_fips$DSCI))
```


```{r, eval = FALSE, include = FALSE}
summary(drought_fips)
```


```{r, eval = FALSE, include = FALSE}
hist(drought_fips$FIPS, main="fips", xlab="percent")
```













```{r}
library(tidyverse)
library(lubridate)
library(tidytuesdayR)
```


```{r}
drought <- read.csv("~/Documents/GitHub/DATA481-Breast_Cancer/Abisai/RData/drought.csv")
head(drought)
tail(drought)
```
```{r}
read_and_prep <- function(file, type){
  read_csv(paste0("2021/2021-07-20/", file)) %>% 
    pivot_longer(cols = None:D4, names_to = "drought_lvl", values_to = type) %>% 
    janitor::clean_names()
  
}

state_area_pct <- read_and_prep("drought_area_pct.csv", "area_pct")
state_area <- read_and_prep("drought_area_total.csv", "area_total")
state_pct_pop <- read_and_prep("drought_pop_pct.csv", "pop_pct")
state_pop <- read_and_prep("drought_pop_total.csv", "pop_total")

all_df <- state_area_pct %>% 
  left_join(state_area) %>% 
  left_join(state_pct_pop) %>% 
  left_join(state_pop) %>% 
  rename(state_abb = state_abbreviation, stat_fmt = statistic_format_id)

all_df %>% 
  write_csv("2021/2021-07-20/drought.csv")

(state_area_pct)
```


```{r}

```












"The U.S. Drought Monitor is jointly produced by the National Drought Mitigation Center at the University of Nebraska-Lincoln, the United States Department of Agriculture, and the National Oceanic and Atmospheric Administration. Map courtesy of NDMC."

Data set on U.S. Droughts from TidyTuesday Github data. Source: U.S. Drought Monitor which describes drought levels from 2001 to 2021 among each state. 

Project: Analyzing drought conditions across the U.S. over ~20 years. State 










































