---
title: "01_make_data_.Rmd"
author: "Abisai Lujan"
date: "2025-01-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(lubridate)
library(tidytuesdayR)
```



**Load drought area percent data**
```{r}
drought_area_pct <- read.csv("drought_area_pct.csv")
drought_area_pct
```
```{r}
str(drought_area_pct)
```
Change `ValidStart` and `ValidEnd` dates to numeric
```{r}
drought_area_pct$ValidStart=as.numeric(gsub("-","",drought_area_pct$ValidStart))
drought_area_pct$ValidEnd=as.numeric(gsub("-","", drought_area_pct$ValidEnd))
```

```{r}
head(drought_area_pct)
```


**Load area total data**

```{r}
drought_area_total <- read.csv("drought_area_total.csv")
head(drought_area_total)
```
Change `ValidStart` and `ValidEnd` to numeric
```{r}
drought_area_total$ValidStart=as.numeric(gsub("-","",drought_area_total$ValidStart))
drought_area_total$ValidEnd=as.numeric(gsub("-","", drought_area_total$ValidEnd))
```
```{r}
head(drought_area_total)
```



**Load drought population percent**

```{r}
drought_pop_pct <- read.csv("drought_pop_pct.csv")
head(drought_pop_pct)
```
Change `ValidStart` and `ValidEnd` dates to numeric
```{r}
drought_pop_pct$ValidStart=as.numeric(gsub("-","",drought_pop_pct$ValidStart))
drought_pop_pct$ValidEnd=as.numeric(gsub("-","",drought_pop_pct$ValidEnd))
```
```{r}
head(drought_pop_pct)
```



**Load drought population total data**
```{r}
drought_pop_total <- read.csv("drought_pop_total.csv")
head(drought_pop_total)
```
Change `ValidStart` and `ValidEnd` dates to numeric 
```{r}
drought_pop_total$ValidStart=as.numeric(gsub("-","",drought_pop_total$ValidStart))
drought_pop_total$ValidEnd=as.numeric(gsub("-","",drought_pop_total$ValidEnd))
```
```{r}
head(drought_pop_total)
```



















```{r}
library(tidyverse)

read_and_prep <- function(file, type){
  read_csv(paste0("2021/2021-07-20/", file)) %>% 
    pivot_longer(cols = None:D4, names_to = "drought_lvl", values_to = type) %>% 
    janitor::clean_names()
  
}

state_area_pct <- read_and_prep("drought_area_pct.csv", "area_pct")
state_area <- read_and_prep("drought_area_total.csv", "area_total")
state_pct_pop <- read_and_prep("drought_pop_pct.csv", "pop_pct")
state_pop <- read_and_prep("drought_pop_total.csv", "pop_total")

all_df <- state_area_pct %>% 
  left_join(state_area) %>% 
  left_join(state_pct_pop) %>% 
  left_join(state_pop) %>% 
  rename(state_abb = state_abbreviation, stat_fmt = statistic_format_id)
```

























































read_and_prep <- function(file, type){
  read_csv(paste0("2021/2021-07-20/", file)) %>% 
    pivot_longer(cols = None:D4, names_to = "drought_lvl", values_to = type) %>% 
    janitor::clean_names()
  
}

state_area_pct <- read_and_prep("drought_area_pct.csv", "area_pct")
state_area <- read_and_prep("drought_area_total.csv", "area_total")
state_pct_pop <- read_and_prep("drought_pop_pct.csv", "pop_pct")
state_pop <- read_and_prep("drought_pop_total.csv", "pop_total")

all_df <- state_area_pct %>% 
  left_join(state_area) %>% 
  left_join(state_pct_pop) %>% 
  left_join(state_pop) %>% 
  rename(state_abb = state_abbreviation, stat_fmt = statistic_format_id)

all_df %>% 
  write_csv("2021/2021-07-20/drought.csv")

(state_area_pct)












"The U.S. Drought Monitor is jointly produced by the National Drought Mitigation Center at the University of Nebraska-Lincoln, the United States Department of Agriculture, and the National Oceanic and Atmospheric Administration. Map courtesy of NDMC."

Data set on U.S. Droughts from TidyTuesday Github data. Source: U.S. Drought Monitor which describes drought levels from 2001 to 2021 among each state. 

Project: Analyzing drought conditions across the U.S. over ~20 years. State 










































