---
title: "Data"
author: "Maanav Jariwala"
date: "2025-01-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(reshape2)
```

```{r}
births <- read_csv("births.csv")
deaths <- read_csv("deaths.csv")
events <- read_csv("events.csv")
```

```{r}
head(births)
head(deaths)
head(events)
```

```{r}
# time series to see the births and deaths over time

ggplot() +
  geom_histogram(data = births, aes(x = year_birth), fill = "blue", alpha = 0.5, binwidth = 10) +
  geom_histogram(data = deaths, aes(x = year_death), fill = "red", alpha = 0.5, binwidth = 10) +
  labs(title = "Distribution of Births and Deaths Over Time",
       x = "Year",
       y = "Count") +
  theme_minimal()
```
```{r}
# Compute correlations
correlation_data <- combined_data %>%
  select(-decade) %>%
  na.omit()

# Melt the correlation matrix
cor_melt <- melt(cor_matrix)

# Plot the heatmap with correlation values
ggplot(cor_melt, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  geom_text(aes(label = sprintf("%.2f", value)), color = "black", size = 4) + # Add correlation values
  labs(title = "Heatmap of Correlations",
       x = "Variables",
       y = "Variables",
       fill = "Correlation") +
  theme_minimal()

```

```{r}
# Add a century column to the events data
events <- events %>% mutate(century = (year %/% 100 + 1))

# Count events by century
events_by_century <- events %>%
  group_by(century) %>%
  summarise(event_count = n())

# Plot events by century
ggplot(events_by_century, aes(x = as.factor(century), y = event_count)) +
  geom_bar(stat = "identity", fill = "green") +
  labs(title = "Number of Events by Century",
       x = "Century",
       y = "Event Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The distribution of births and deaths reveals historical trends, such as potential population growth periods or impacts of events like wars and pandemics.

The number of events by century suggests increased human activity or record-keeping improvements. A study could explore the drivers behind these trends.

The correlation heatmap provides insights into how events, births, and deaths are interrelated, for a starting point for deeper historical or sociological analysis.

