---
title: "droughtEDA"
author: "Samuel Carlton"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
drought_pct <- read.csv("~/Downloads/drought_area_pct.csv")
head(drought_pct)
```

```{r}
#unique(drought_pct$StateAbbreviation)
```


```{r}
library(dplyr)

drought_pct <- drought_pct %>%
  mutate(Moderate_Drought = None + D0 + D1,
         Severe_Drought = D2 + D3 + D4)

head(drought_pct)

```



# Cali, Nevada, NM, Utah, Arizona
```{r}
# Filter dataset for selected states
drought_pct_filtered <- drought_pct %>%
  filter(StateAbbreviation %in% c("CA", "NV", "NM", "UT", "AZ"))

# Display the first few rows
head(drought_pct_filtered)
```

```{r}
# Select only the necessary columns
drought_pct_filtered <- drought_pct_filtered %>%
  select(MapDate, StateAbbreviation, ValidStart, Moderate_Drought, Severe_Drought)

# Display the first few rows
head(drought_pct_filtered)
```

```{r}
# Load necessary library
library(dplyr)
library(lubridate)

# Convert MapDate column to a proper Date format
FinalDrought <- drought_pct_filtered %>%
  mutate(MapDate = ymd(MapDate))

# Display the first few rows to verify
head(FinalDrought)

```

```{r}
library(ggplot2)

# Line graph for Moderate and Severe Drought
ggplot(FinalDrought, aes(x = MapDate)) +
  geom_line(aes(y = Moderate_Drought, color = "Moderate Drought"), size = 1) +
  geom_line(aes(y = Severe_Drought, color = "Severe Drought"), size = 1) +
  facet_wrap(~StateAbbreviation, scales = "free_y") +  # Separate plots for each state
  labs(title = "Drought Trends Over Time",
       x = "Date",
       y = "Drought Percentage",
       color = "Drought Severity") +
  theme_minimal()
```
```{r}
yearly_avg_severe <- FinalDrought %>%
  mutate(Year = format(MapDate, "%Y")) %>%
  group_by(Year, StateAbbreviation) %>%
  summarize(Avg_Severe = mean(Severe_Drought, na.rm = TRUE))

ggplot(yearly_avg_severe, aes(x = Year, y = Avg_Severe, color = StateAbbreviation, group = StateAbbreviation)) +
  geom_line() +
  labs(title = "Yearly Trends in Severe Drought",
       x = "Year",
       y = "Avg Severe Drought (%)") +
  theme_minimal()

```

```{r}
ggplot(FinalDrought, aes(x = StateAbbreviation, y = Moderate_Drought, fill = StateAbbreviation)) +
  geom_boxplot() +
  labs(title = "Distribution of Moderate Drought by State", y = "Moderate Drought (%)") +
  theme_minimal()
  
ggplot(FinalDrought, aes(x = StateAbbreviation, y = Severe_Drought, fill = StateAbbreviation)) +
  geom_boxplot() +
  labs(title = "Distribution of Severe Drought by State", y = "Severe Drought (%)") +
  theme_minimal()

```


```{r}
library(reshape2)

FinalDrought_melted <- melt(FinalDrought, id.vars = c("MapDate", "StateAbbreviation"), 
                            measure.vars = c("Moderate_Drought", "Severe_Drought"))

ggplot(FinalDrought_melted, aes(x = MapDate, y = StateAbbreviation, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "yellow", high = "red") +
  facet_wrap(~variable) +
  labs(title = "Drought Severity Heatmap",
       x = "Date",
       y = "State",
       fill = "Drought Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
FinalDrought$Month <- format(FinalDrought$MapDate, "%m")

avg_monthly_severe <- FinalDrought %>%
  group_by(StateAbbreviation, Month) %>%
  summarize(Avg_Severe = mean(Severe_Drought, na.rm = TRUE))

ggplot(avg_monthly_severe, aes(x = Month, y = Avg_Severe, color = StateAbbreviation, group = StateAbbreviation)) +
  geom_line() +
  labs(title = "Seasonal Trends in Severe Drought",
       x = "Month",
       y = "Avg Severe Drought (%)") +
  theme_minimal()

```
# Identify Peak Drought Nmonths for Each State
```{r}
peak_months <- FinalDrought %>%
  group_by(StateAbbreviation, Month) %>%
  summarize(Avg_Severe = mean(Severe_Drought, na.rm = TRUE)) %>%
  arrange(StateAbbreviation, desc(Avg_Severe)) %>%
  slice(1)  # Get the peak month for each state

print(peak_months)

```
# Compare Seasonality Over Different Time Periods
```{r}
FinalDrought$Decade <- ifelse(FinalDrought$MapDate < "2010-01-01", "2000-2010", "2011-2023")

seasonal_trends_by_decade <- FinalDrought %>%
  group_by(Decade, Month, StateAbbreviation) %>%
  summarize(Avg_Severe = mean(Severe_Drought, na.rm = TRUE))

ggplot(seasonal_trends_by_decade, aes(x = Month, y = Avg_Severe, color = StateAbbreviation, group = StateAbbreviation)) +
  geom_line() +
  facet_wrap(~Decade) +
  labs(title = "Severe Drought Seasonality by Decade",
       x = "Month",
       y = "Avg Severe Drought (%)") +
  theme_minimal()

```

# Compare Drought Recovery Speed Between States
```{r}
state_recovery_summary <- FinalDrought %>%
  group_by(StateAbbreviation) %>%
  summarize(
    Min_Recovery = min(Drought_Recovery, na.rm = TRUE),
    Q1_Recovery = quantile(Drought_Recovery, 0.25, na.rm = TRUE),
    Median_Recovery = median(Drought_Recovery, na.rm = TRUE),
    Mean_Recovery = mean(Drought_Recovery, na.rm = TRUE),
    Q3_Recovery = quantile(Drought_Recovery, 0.75, na.rm = TRUE),
    Max_Recovery = max(Drought_Recovery, na.rm = TRUE),
    Count_Observations = n()
  )

print(state_recovery_summary)
```
Why are Q1, Median, and Q3 almost always 0? This suggests most months did not experience a change in severe drought conditions. 

# Check distribution of Drought_Recovery values to confirm if extreme values occure in only a few months.
```{r}
ggplot(FinalDrought, aes(x = Drought_Recovery)) +
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7) +
  facet_wrap(~StateAbbreviation, scales = "free_y") +
  labs(title = "Distribution of Drought Recovery Rates by State",
       x = "Drought Recovery (%)",
       y = "Count") +
  theme_minimal()

```
Most values clustered at 0, so may be better to focus only on periods of significant recovery.

# Identify States with the Fastest Recovery Rates
```{r}
fast_recovery_events <- FinalDrought %>%
  filter(Drought_Recovery < -30) %>%  # Only major drops
  arrange(MapDate)

print(fast_recovery_events)

```
Why/how did these recovery events occur?


# Compare Recovery Rates Over Time
```{r}
FinalDrought$Decade <- ifelse(FinalDrought$MapDate < "2010-01-01", "2000-2010", "2011-2023")

recovery_trends <- FinalDrought %>%
  group_by(Decade, StateAbbreviation) %>%
  summarize(Avg_Recovery = mean(Drought_Recovery, na.rm = TRUE))

ggplot(recovery_trends, aes(x = Decade, y = Avg_Recovery, fill = StateAbbreviation)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Drought Recovery Speed Over Time",
       x = "Decade",
       y = "Avg Recovery (%)") +
  theme_minimal()

```
Recovery seems faster in the most recent decade. Why?

# Find States with the Most Extreme Drought Swings
```{r}
volatility <- FinalDrought %>%
  group_by(StateAbbreviation) %>%
  summarize(Std_Dev_Recovery = sd(Drought_Recovery, na.rm = TRUE))

print(volatility)

```

